### Dataset features

Both the Fatal Encounters project and the Washington Post are open-source, "unofficial" datasets.  Their data are not drawn from mandatory government reporting on deaths caused by law enforcement officers because there is no official national data program designed for this purpose.  The datasets are based instead on cases known/discovered by the two project teams, primarily through online searches.  You can find more details about the methodology used by each project on their websites:  [Fatal Encounters](https://fatalencounters.org/methodology/) and [Washington Post](https://www.washingtonpost.com/investigations/2022/12/05/washington-post-fatal-police-shootings-methodology/).  

#### Open-source vs. Official

In theory, deaths caused by police should also be captured by several official data collection programs:  

* the [National Vital Statistics System](https://www.cdc.gov/violenceprevention/datasources/nvdrs/index.html), which collects death certificates

* the [Arrest Related Death](https://bjs.ojp.gov/data-collection/arrest-related-deaths-ard) program, run by the Bureau of Justice Statistics (BJS), suspended in 2014, current status unknown.

* the [Supplementary/Expanded Homicide Reports](https://ucr.fbi.gov/crime-in-the-u.s/2019/crime-in-the-u.s.-2019/tables/expanded-homicide-data-table-14.xls) program, run by the the Federal Bureau of Investigation (FBI), based on the Uniform Crime Reporting program.

* the [National Violent Deaths Reporting System](https://www.cdc.gov/violenceprevention/datasources/nvdrs/index.html), a newer program run by the Centers for Disease Control and Prevention (CDC), all states reporting as of 2018.


Studies have repeatedly shown, however, that these official programs consistently fail to identify about half of the people killed by police.  The most recent study to have replicated this finding was conducted by the [Institute for Health Metrics and Evaluation](https://www.healthdata.org/about) in 2021, and published in the [Lancet](https://www.thelancet.com/).  You can access the study [here](https://www.thelancet.com/journals/lancet/article/PIIS0140-6736(21)01609-3/fulltext), and less technical summaries [can be found in the news media](https://www.cnn.com/2021/09/30/us/study-police-deaths-undercount/index.html).  Other studies show similar rates of undercounting in the Arrest Related Death and Supplementary Homicide Reporting systems, one example is [here](https://www.ojp.gov/library/publications/assessment-coverage-arrest-related-deaths-program).  

The one exception to this is the National Violent Deaths Reporting system, where a [study from 2015](https://ajph.aphapublications.org/doi/full/10.2105/AJPH.2018.304904) found that it captured almost all of the people shot and killed by police in the 27 states that were participating in the program at the time.  But there has been no update since the state coverage reached all 50 states in 2018, and there has been no assessment of its coverage of deaths due to causes other than gunshot.  Perhaps most importantly, the study simply showed that this dataset had comparable coverage to the open-source datasets, not that it was superior.

Research has also shown that the methodology used by the open-source data sets is effective:

  "Findings indicate that the open-source methodology alone identifies the majority of law enforcement homicides, but agency surveys aid in identifying deaths by other causes (e.g., accidents, suicides, and natural causes)."  [BJS 2019](https://www.ojp.gov/library/publications/arrest-related-deaths-program-pilot-study-redesigned-survey-methodology)

_____

> At this point, it is widely acknowleged that the open-source datasets have the most complete coverage of these cases.  That is why we use them here.

_____

The other key benefit to the Fatal Encounters data in particular is that every case includes a link to documentation that verifies the case and serves as a starting place for additional research.  This allows us to identify the law enforcement agency(ies) involved, provides more context and sometimes reveals conflicting narratives as to the events.  All of this helps to remind us that each case is a person, not just a statistic.

That said, it is important to keep in mind that these open-source datasets, too, are not perfect.   They will miss cases that have not generated a digital signature that can be found using their search methodologies.  So it would be accurate to say that our open-source data provide a lower bound estimate of the number of people killed by police; the true value is likely higher, but it can not be lower.

#### Dataset differences

Fatal Encounters includes more cases than the Washington Post, because the Post dataset is restricted to fatal shootings.  

____

> Among the cases missing from the Washington Post data are Manny Ellis (in WA) and George Floyd (at the national level), because their deaths were caused by asphyxiation, not gunshots.

____

The Washington Post also excludes deaths that occur during police vehicular pursuits.  As can be seen in this report, police pursuits are the second leading cause of death in police-civilian encounters, so their inclusion in Fatal Encounters is important.  You can find more details on pursuit fatalities here in WA State in [this report](https://rpubs.com/moxbox/wa_pursuits).

The total number of cases in the Fatal Encounters dataset for WA State since
`r params$from` is `r nrow(merged_data)`.  These are classified as follows:

```{r tabAllCases}

tab <- merged_data %>%
  group_by(circumstances) %>%
  summarize(Number = n(),
            Percent = round(100*Number/nrow(merged_data), 1)
  ) %>%
  arrange(desc(Number)) %>%
  bind_rows(data.frame(circumstances ="Total", 
                       Number = sum(.$Number), 
                       Percent = sum(.$Percent))) 

tab %>%
  kable(caption = "All Fatal Encounters dataset cases, broken down by type",
        col.names = c("Type", "Number", "Percent")) %>%
  kable_styling(bootstrap_options = c("striped","hover")) %>%
  row_spec(dim(tab)[1], bold = T)  %>%
  add_footnote(label = "Percents may not sum to 100 due to rounding",
               notation = "symbol")
```
  
### Suicides and "Medical Emergencies" excluded

The Fatal Encounters data includes some cases classified as "suicides" that occur during an encounter with police.  These are not cases of "suicide by cop", where the victim appears to have provoked the police in order to be shot by police.  They are cases where the victim shoots or otherwise kills themself.  **We have excluded these from the analyses in this report, because the report focuses on persons killed by police.**  

However, we acknowledge that these cases warrant further assessment.  The classification as "suicide" is not definitive, it is typically based on the narrative provided by law enforcement.  And even if these are suicides, it is possible that the outcome might have been different if the case had been handled differently, for example, by calling in a crisis intervention team.

The excluded case counts and fraction of total cases by year is shown in the table below.

```{r tabSuicide}

tab <- merged_data %>%
  group_by(year, homicide) %>%
  summarize(Number = n()) %>%
  pivot_wider(id_cols = year,
              names_from = homicide,
              names_prefix = "N",
              values_from = Number) %>%
  replace_na(list(N0=0)) %>%
  mutate(Percent = round(100*N0/(N0 + N1), 1),
         year = as.character(year)) %>%
  select(year, Number=N0, Percent) %>%
  bind_rows(data.frame(year ="Total", 
                       Number = sum(.$Number), 
                       Percent = round(100*sum(.$Number)/nrow(merged_data), 1)))

tab %>%
  kable(caption = "Cases labeled 'Suicide' in Fatal Encounters dataset by year",
        col.names = c("Type", "Number", "Percent of all cases")) %>%
  kable_styling(bootstrap_options = c("striped","hover"),
                full_width = F) %>%
  row_spec(dim(tab)[1], bold = T)  %>%
  add_footnote(label = "These cases were excluded from analyses in this report",
               notation = "symbol")
```
There are also a couple of cases where the cause of death is coded as "Medical emergency".  We exclude these cases from this report as well, as they do not meet the definition of "homicide."  But we note that in all of these cases, there was substantial interaction with the police in the period preceding the death, and the person died in police custody.  The cases in WA both involved a lengthy vehicular pursuit, and in the national level dataset, many of these cases involved the victim being restrained and injected with Ketamine by emergency responders.

___


### Homicide case comparison

Restricting the Fatal Encounters dataset to the homicide cases, it still has 
`r paste0(round(100*prop.table(table(is.na(homicides$wapoID)))[[2]]), "%")`
more cases than found in the Post dataset.

```{r difftab}
tab <- homicides %>%
  mutate(in.wapo = ifelse(is.na(wapoID), 
                          "In FE only", 
                          "In both FE and WaPo")) %>%
  group_by(in.wapo) %>%
  summarize(Number = n(),
            Percent = round(100*Number/nrow(homicides), 1)
  ) %>%
  bind_rows(data.frame(in.wapo ="Total", 
                       Number = sum(.$Number), 
                       Percent = sum(.$Percent))) 

tab %>%
  kable(caption = "Cases found in each dataset",
        col.names = c("Case is:", "Number", "Percent")) %>%
  kable_styling(bootstrap_options = c("striped","hover")) %>%
  row_spec(row=dim(tab)[1], bold = T) %>%
  add_footnote(label = "Percents may not sum to 100 due to rounding",
               notation = "symbol")

```

___

#### Missing cases in WaPo {.tabset}

##### By cause of death


```{r datadiff}
tab <- homicides %>%
  filter(is.na(wapoID)) %>%
  group_by(cod) %>%
  summarize(Number = n()
  ) %>%
  arrange(desc(Number)) %>%
  bind_rows(data.frame(cod ="Total", 
                       Number = sum(.$Number))) 

pct.shot.missing <- tab$Number[tab$cod=="Gunshot"]/tab$Number[tab$cod=="Total"]
```

Among the cases missing from the Post dataset, 
`r tab$Number[tab$cod=="Gunshot"]` (`r scales::percent(pct.shot.missing)`)
are identified as fatal shootings in the Fatal Encounters dataset.

```{r datadifftab}
tab %>%
  kable(caption = "Cause of death for cases missing from WaPo dataset",
        col.names = c("Cause:", "Number")) %>%
  kable_styling(bootstrap_options = c("striped","hover")) %>%
  row_spec(row=c(which(tab$cod=="Gunshot"),
                 dim(tab)[1]), bold = T)
```

____

##### By name, for fatal shootings

Below is the list of cases identified as gunshot fatalities in Fatal Encounters that are missing from the Washington Post dataset.

```{r casesdiff}

homicides %>%
  filter(is.na(wapoID) & cod == "Gunshot") %>%
  select(name, date, city, county, agency, url_click) %>%
  DT::datatable(rownames = F,
                caption = "Gunshot cases missing from WaPo dataset",
                escape = FALSE)

```

____

